\sysfull \\ OR \\ \sysfullalt

### 2017-07-25 paper/background.tex
Still working on a background section

The biggest technical problem with energy provided from the ambient to battery-less platforms is that harvested energy is not stable and difficult to predict accurately. Combined with the fact that energy supply is small, there is little leeway to store enough of energy to guarantee prolonged periods of computation. Energy breaks happen every hundred of milliseconds, refer again to e.g.~\cite[Fig. 1]{mementos}. This causes each program running on such platform to restart. Solution to the problem is to divide the program into smaller parts that guarantee execution within a discharge region and track the state of the program in-between execution times.

@@

Naturally, content of a volatile memory (i.e. run-time stack, global and local variables) of the microcontroller and program registers are erased after each power failure. Only what resides in a non-volatile memory, e.g. FRAM, persists after system reboot. A solution to sustain the progress of the program from the last power failure, instead from the beginning (the \emph{Sisyphus} effect~\cite[Sec. 2]{mementos})is a periodic checkpointing of a program state~\cite{mementos,hibernusplusplus,quickrecall,idetic}. The periodicity, and a position of a checkpoint, is the core of the contribution behind each of the proposed checkpoining solution.

@@

Nevertheless, regular checkpointing of program state to a non-volatile memory still does not guarantee consistency. This phenomenon has been discussed extensively in e.g.~\cite{chain,alpaca}. For the completeness of this exposition we will provide our own example demonstrating this problem. For this please refer to Fig.~\ref{fig:code_demo_incosistency} and compare with~\cite[Sec 2.3]{alpaca},~\cite[Sec. 2.1]{chain}. We take an example of a fast procedure to calculate cyclic redundancy check (CRC) from~\cite{hicks_mibench2_2016}. Therein a CRC must be calculated for $\texttt{nBytes}$-long message $\texttt{msg}$ with specified initial reminder $\texttt{Rmnd}$ and constant, $\texttt{Rmnd}$ variable type-dependent, $\texttt{Wdth}$. Despite regular, fixed position checkpoints (line 3 in Fig.~\ref{fig:crc_example}) of all variables which are declared as non-volatile (denoted as \texttt{NV}) variable $\texttt{data}$ is calculated incorrectly. This is because of a write after read (WAR) hazard of variable $\texttt{Rmnd}$. WAR hazard is caused by consecutive write and read in-between two consecutive checkpoints using $\texttt{checkpoint()}$ function: $\texttt{data}$ variable is written \emph{before} (line 10 in Fig.~\ref{fig:crc_example_war}) $\texttt{Rmnd}$ is read (line 6 in Fig.~\ref{fig:crc_example_war}), or more specifically---updated before a new execution loop.

We assume that we operate on the microcontroller (more generally, computing hardware) equipped with a mix of volatile and non-volatile memory, such as FLASH, Ferroelectric RAM or Magnetoresistive RAM. \sys does not assume to rely on specific memory type (either volatile or non-volatile), in contrary to earlier works targeting non-volatile processors~\cite{su_date_2017,ratchet,quickrecall,nvp}. Actually, \sys will exploit the properties of both memory types to maximize performance metrics of the program being executed. We make \sys able to I/O operations, however we do not enable the scheduling of operations (which \emph{nota bene} forms a foundation for a simple runtime). All in all, our hardware model follows in assumptions the two most relevant intermittently-powered execution environments~\cite{alpaca,chain}.

@@

There is increasing amount of work that deals with protecting memory consistency. Table~\ref{table:chechpoint_comparison} provides a summary of various costs associated with most relevant task-based/checkpointing solutions. This table generalizes discussion from~\cite[Sec. 2.4]{alpaca}.

@@

The energy cost of a single checkpoint is $w\text{CP(SZ+NV)}$, where $w$ is the cost of a single word copy into a non-volatile memory. Decrease of CP decreases the energy cost associated with each state preservation method and reduces operation cost which in most simplistic terms is defined as $\mathcal{O}(\text{CP})+\mathcal{O}(\text{PL})$, where PL is the cost of program execution without checkpointing. On the other hand, this increases task/checkpoint re-execution time.

Moreover, each battery-less platform has different storage sizes, refer to Table~\ref{table:capacitor}, which causes different charge/discharge periods. This result in different execution times. In the worst case the size of the capacitor might be too small to execute one task written by the programmer.

>>>>

The biggest technical problem with energy provided from the ambient to battery-less platforms is that harvested energy is not stable and difficult to predict accurately. Combined with the fact that energy supply is small, there is little leeway to store enough of energy to guarantee prolonged periods of computation. Energy breaks happen every hundred of milliseconds, refer again to e.g.~\cite[Fig. 1]{mementos}. This causes each program running on such platform to restart and requires a fundamentally new model for program execution:  \emph{intermittent} (discontinuous) execution model. In the conventional (or \emph{continuous}) execution model a power failure is seen as an \emph{exception} that may reset the progress of a program to its start. Whereas, in an intermittent execution model a power failure is regarded as a temporary \emph{pause} to the execution consequently prolonging execution time. A conceptual implementation of intermittent execution model is to divide the program into smaller parts that guarantee execution within a discharge region and track the state of the program in-between execution times.

Before proceeding further with the background we need to make certain assumptions about the underlying hardware. We assume that we operate on the microcontroller (more generally, computing hardware) equipped with a mix of volatile and non-volatile memory, such as FLASH, Ferroelectric RAM or Magnetoresistive RAM. \sys does not assume to rely on specific memory type (either volatile or non-volatile), in contrary to earlier works targeting non-volatile processors~\cite{su_date_2017,ratchet,quickrecall,nvp}. Actually, \sys will exploit the properties of both memory types to maximize performance metrics of the program being executed. We make \sys able to I/O operations, however we do not enable the scheduling of operations (which \emph{nota bene} forms a foundation for a simple runtime). All in all, our hardware model follows in assumptions the two most relevant intermittently-powered execution environments~\cite{alpaca,chain}.

@@

Naturally, content of a volatile memory (i.e. run-time stack, global and local variables) of the microcontroller and program registers are erased after each power failure. Only what resides in a non-volatile memory, e.g. FRAM, persists after system reboot. A solution to sustain the progress of the program from the last power failure, instead from the beginning (the \emph{Sisyphus} effect~\cite[Sec. 2]{mementos})is a periodic checkpointing of a program state~\cite{mementos,hibernusplusplus,quickrecall,idetic}. The periodicity, and a position of a checkpoint, as well as the way to combat the \emph{program state consistency} phenomenon (discussed extensively in e.g.~\cite{chain,alpaca}), is the core of the contribution behind each of the proposed checkpoining solution. For the completeness of this exposition we will provide our own example demonstrating this phenomenon, viz. please refer to Fig.~\ref{fig:code_demo_incosistency} and compare with~\cite[Sec 2.3]{alpaca},~\cite[Sec. 2.1]{chain}.

@@

Let us take an example of a fast procedure to calculate cyclic redundancy check (CRC) from~\cite{hicks_mibench2_2016}. Therein a CRC must be calculated for $\texttt{nBytes}$-long message $\texttt{msg}$ with specified initial reminder $\texttt{Rmnd}$ and constant, $\texttt{Rmnd}$ variable type-dependent, $\texttt{Wdth}$. Despite regular, fixed position checkpoints (line 3 in Fig.~\ref{fig:crc_example}) of all variables which are declared as non-volatile (denoted as \texttt{NV}) variable $\texttt{data}$ is calculated incorrectly. This is because of a write after read (WAR) hazard of variable $\texttt{Rmnd}$. WAR hazard is caused by consecutive write and read in-between two consecutive checkpoints using $\texttt{checkpoint()}$ function: $\texttt{data}$ variable is written \emph{before} (line 10 in Fig.~\ref{fig:crc_example_war}) $\texttt{Rmnd}$ is read (line 6 in Fig.~\ref{fig:crc_example_war}), or more specifically---updated before a new execution loop.

@@

Generally we can classify the intermittent execution model into:

\subsubsection{Sequential Execution Model}: Under this model a program is seen as one big idempotent operational region that has one common context. Generally, the progress of the program is saved and updated by means of checkpointing, refer to Section~\ref{sec:background_consistency}. Normally, the sequential model relies on a hardware assistant to measure the voltage level in the energy reservoir to place a checkpoint~\cite{mementos,mottola2017harvos,hibernus}. The benefit of this model is that it does not require code modification by a programmer. However, it has a number of drawbacks: (i) It suffers from significant overhead \cite{chain}; (ii) the programmer should not access the non-volatile memory to guarantee the consistency of the memory~\cite{}; and (iii) it restricts the intermittently-powered devices to run only a single application.

\subsubsection{Modular Intermittent Execution Model}: At the heart of this model is the concept of an \emph{idempotent task}. The idempotent task is a C language function that does not have arguments and does not return a value. This task uses a well defined interface to interact with the non-volatile memory. Therefore, it tolerates arbitrary number of power interrupts. This model, generally, produces less overhead~\cite{chain} and allows multiple applications to run on an intermittently-powered device by interleaving their tasks~\cite{}. However, it obviously requires code modification. For example, if an algorithm is written according to the continuous execution model it has to be split, by a programmer, into small tasks to run under the modular intermittent execution model.

Both modular and intermittent models are present in the literature. Table~\ref{table:chechpoint_comparison} provides a summary of various costs associated with most relevant task-based/checkpointing solutions. This table generalizes discussion from~\cite[Sec. 2.4]{alpaca}.

@@

\textbf{Cost of a checkpoint:} Whatever the approach, the energy cost of a single checkpoint is $w\text{CP(SZ+NV)}$, where $w$ is the cost of a single word copy into a non-volatile memory. Decrease of CP decreases the energy cost associated with each state preservation method and reduces operation cost which in most simplistic terms is defined as $\mathcal{O}(\text{CP})+\mathcal{O}(\text{PL})$, where PL is the cost of program execution without checkpointing. On the other hand, this increases task/checkpoint re-execution time.

\textbf{Cost of a fixed task size:} A modular intermittent execution model completely depends on a programmer's estimation which mostly results in a sub-optimal code division. Moreover, this model can only consider a single hardware configuration and it does not take environment changing into considerations. Each battery-less platform has different storage sizes, refer to Table~\ref{table:capacitor}, which causes different charge/discharge periods. This result in different execution times and in the worst case the size of the capacitor might be too small to execute one task written by the programmer.

### 2017-07-25 paper/paper.tex
title suggestion
Still working on a background section

\newcommand{\TODO}[1]{{\bf \em \textcolor{red}{TODO: #1}\xspace}}
\newcommand{\todo}[2]{\textcolor{red}{$\bullet$ \textbf{Task:} #1; \textbf{Responsible:} #2}}
\newcommand{\sys}{VIPE\xspace}
\newcommand{\sysfull}{Virtualized Intermittently-Powered Executor\xspace}

@@

\title{VIPE: Virtualized Intermittently-Powered Executor}

@@

\section{Memory Virtualization}
\label{sec:memory_virtulaization}
\input{paging}
\begin{figure}
	\centering
	%\includegraphics[width=0.25\columnwidth]{figures/}
	\caption{Memory virtualization architecture.}
	\label{fig:}
\end{figure}

@@

\section{Old Introduction}
	Intermittently powered devices (IPDs) are battery-less devices that utilize the ambient energy to sense, compute and, communicate. For example, Wireless Identification and Sensing platform WISP \cite{wisp} uses the RF signal power to drive its computation and communication. Because of the reliance of TPDs on intermittent power supply, the harvested energy when it is available, their programs are susceptible to a very frequent power interrupts, in the order of tens of millisconds~\cite{}. Therefore, these devices require a different software execution model that complies with the nature of a discontinuous power supply.

	The intermittent (discontinuous) execution model defines a program execution as cumulative discrete process. The main difference between the intermittent and the conventional (or continuous) execution models is that, a power failure is seen by the continuous model as an \emph{exception} that may reset the progress of a program to its beginning. Whereas, in intermittent execution a power failure is regarded as a temporary \emph{pause} to the execution that may result in some progress degradation. Generally we can classify the intermittent execution model into: 
	\begin{itemize}
		\item \emph{Sequential Execution Model}:
			Under this model a program is seen as one big idempotent operational region that has one common context. Generally, The progress of the program is saved and updated by means of checkpointing---where all the program context (e.g. CPU registers, the stack and the global variables) is saved to a non-volatile memory. Normally, the sequential model relies on a hardware assistant to measure the voltage level in the energy reservoir to place a checkpoint~\cite{mementos, harvOS, hibernus}. The benefit of this model is that it does not require code modification by a programmer. However, it has a number of drawbacks: (i) It suffers from significant overhead \cite{chain}; (ii) the programmer should not access the non-volatile memory to guarantee the consistency of the memory~\cite{xxxx}; and (iii) it restricts the IPDs to run only a single application~\cite{inos}. 
		\item \emph{Modular Intermittent Execution Model}:
			At the heart of this model is the concept of an idempotent task. The idempotent task is C function that does not have arguments and does not return a value. This task uses a well defined interface to interact with the non-volatile memory. Therefore, it tolerates arbitrary number of power interrupts. This model, generally, produces less overhead~\cite{chain} and allows multiple applications to run on an IPD by interleaving their tasks~\cite{inos}. However, it obviously requires code modification---for example, if an algorithm is written according to the continuous execution model it has to be splitted, by a programmer, into small tasks to run under the modular intermittent execution model.
	\end{itemize}

Despite that the superiority of the Modular Intermittent Execution model, it is still a static approach that completely depends on a programmer's estimation which is mostly result in a sub-optimal code devision. Moreover, this model can only consider a single hardware configuration and it does not take environment changing into considerations.

@@

\subsection{Why not Adapting Current Approaches}
Here we highlight the challenges of adapting the state-of-the-art proposed methods to enable the execution of dynamic task sizes. 
%
%	\begin{figure}[t]
%	    \centering
%	         \subfigure[Chain: Sequential task execution flow control.]{\includegraphics[width=0.35\columnwidth]{figures/dynamic-chain.pdf} \label{fig:DynamicChainSeq} } 
%	         \subfigure[Chain: Random task execution flow control.]{\includegraphics[width=0.61\columnwidth]{figures/dynamic-chain-2.pdf} \label{fig:DynamicChainRan}}
%	    \caption{Task flow control of Chain}
%	    \label{fig:DynamicChain}
%	\end{figure}

>>>>

\newcommand{\TODO}[1]{{\bf \em \textcolor{red}{TODO: #1}\xspace}}
\newcommand{\todo}[2]{\textcolor{red}{$\bullet$ \textbf{Task:} #1; \textbf{Responsible:} #2}}
\newcommand{\sys}{VIPE\xspace}
\newcommand{\sysfull}{Virtualized Intermittently-Powered Executor\xspace}
\newcommand{\sysfullalt}{Virtualized Intermittently-Powered Execution Runtime\xspace}

@@

\title{\sysfull \\ OR \\ \sysfullalt}

@@

\section{Compiler Support for Automatic Task Generation}
\label{sec:compiler}

\input{compiler}

\section{Memory Virtualization}
\label{sec:memory_virtulaization}

\input{paging}

\begin{figure}
	\centering
	%\includegraphics[width=0.25\columnwidth]{figures/}
	\caption{Memory virtualization architecture.}
	\label{fig:}
\end{figure}

@@

\section{Old Introduction}

@@

\subsection{Why not Adapting Current Approaches}

Here we highlight the challenges of adapting the state-of-the-art proposed methods to enable the execution of dynamic task sizes. 
%
%	\begin{figure}[t]
%	    \centering
%	         \subfigure[Chain: Sequential task execution flow control.]{\includegraphics[width=0.35\columnwidth]{figures/dynamic-chain.pdf} \label{fig:DynamicChainSeq} } 
%	         \subfigure[Chain: Random task execution flow control.]{\includegraphics[width=0.61\columnwidth]{figures/dynamic-chain-2.pdf} \label{fig:DynamicChainRan}}
%	    \caption{Task flow control of Chain}
%	    \label{fig:DynamicChain}
%	\end{figure}
